---
title: "Postop Cytokines over Time"
author: "Gregg Griffenhagen"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
library(tidyverse)
library(emmeans)
library(lme4)
library(lmerTest)
library(ggplot2)
library(ggpubr)
library(kableExtra)
data <- read.csv('postop_cytok_over_time.csv')
data$surgery <- as.factor(data$surgery)
data$horse <- as.factor(data$horse)
```

## Quick Data Summary

```{r, warning=FALSE}
data %>% group_by(time, surgery) %>%
  summarise(across(c(il.18, il.2),
                   c(mean = ~mean(.x), se = ~sd(.x)/sum(!is.na(.x)),
                     min = ~min(.x), max = ~max(.x)))) %>%
  kable() %>%
  kable_styling(font_size = 8, position = 'left', latex_options = 'hold_position')
data %>% group_by(time, surgery) %>%
  summarise(across(c(ip.10, il.10),
                   c(mean = ~mean(.x), se = ~sd(.x)/sum(!is.na(.x)),
                     min = ~min(.x), max = ~max(.x)))) %>%
  kable() %>%
  kable_styling(font_size = 8, position = 'left', latex_options = 'hold_position')
data %>% group_by(time, surgery) %>%
  summarise(across(c(temp, wbc),
                   c(mean = ~mean(.x, na.rm=T), se = ~sd(.x, na.rm=T)/sum(!is.na(.x)),
                     min = ~min(.x, na.rm=T), max = ~max(.x, na.rm=T)))) %>%
  kable() %>%
  kable_styling(font_size = 8, position = 'left', latex_options = 'hold_position')
data %>% group_by(time, surgery) %>%
  summarise(across(c(neut, lymph),
                   c(mean = ~mean(.x, na.rm=T), se = ~sd(.x, na.rm=T)/sum(!is.na(.x)),
                     min = ~min(.x, na.rm=T), max = ~max(.x, na.rm=T)))) %>%
  kable() %>%
  kable_styling(font_size = 8, position = 'left', latex_options = 'hold_position')
data %>% group_by(time, surgery) %>%
  summarise(across(c(fibr),
                   c(mean = ~mean(.x, na.rm=T), se = ~sd(.x, na.rm=T)/sum(!is.na(.x)),
                     min = ~min(.x, na.rm=T), max = ~max(.x, na.rm=T)))) %>%
  kable() %>%
  kable_styling(font_size = 8, position = 'left', latex_options = 'hold_position')
```

# Start with the plotting full cytokine data set

```{r, echo=FALSE}
p1 <- ggplot(data = data, aes(x = time, y = il.18, group = horse, color = surgery))
p1 <- p1 + geom_point() +
  geom_line()

p2 <- ggplot(data = data, aes(x = time, y = il.10, group = horse, color = surgery))
p2 <- p2 + geom_point() +
  geom_line()

p3 <- ggplot(data = data, aes(x = time, y = il.2, group = horse, color = surgery))
p3 <- p3 + geom_point() +
  geom_line()

p4 <- ggplot(data = data, aes(x = time, y = ip.10, group = horse, color = surgery))
p4 <- p4 + geom_point() +
  geom_line()

p5 <- ggarrange(p1, p2, p3, p4, ncol = 2, nrow = 2, common.legend = TRUE, legend = "bottom")

p5
```

## Remove the obvious outlier at the top and recheck plots

```{r, echo=FALSE}
data2 <- data %>% filter(horse != 13) #removes the outlier manually

p1_2 <- ggplot(data = data2, aes(x = time, y = il.18, group = horse, color = surgery))
p1_2 <- p1_2 + geom_point() +
  geom_line()

p2_2 <- ggplot(data = data2, aes(x = time, y = il.10, group = horse, color = surgery))
p2_2 <- p2_2 + geom_point() +
  geom_line()

p3_2 <- ggplot(data = data2, aes(x = time, y = il.2, group = horse, color = surgery))
p3_2 <- p3_2 + geom_point() +
  geom_line()

p4_2 <- ggplot(data = data2, aes(x = time, y = ip.10, group = horse, color = surgery))
p4_2 <- p4_2 + geom_point() +
  geom_line()

p5_2 <- ggarrange(p1_2, p2_2, p3_2, p4_2, ncol = 2, nrow = 2, common.legend = TRUE, legend = "bottom")

p5_2
```

## The 3 interleukins still demonstrate a single horse as an outlier (although not as bad)
## Let's remove that one too

```{r, echo=FALSE}
data3 <- data2 %>% filter(horse != 2) # again remove the outlier manually

p1_3 <- ggplot(data = data3, aes(x = time, y = il.18, group = horse, color = surgery))
p1_3 <- p1_3 + geom_point() +
  geom_line()

#FOCUS ON THIS ONE  - seems to be the only one with differences
p2_3 <- ggplot(data = data3, aes(x = time, y = il.10, group = horse, color = surgery))
p2_3 <- p2_3 + geom_point() +
  geom_line()

p3_3 <- ggplot(data = data3, aes(x = time, y = il.2, group = horse, color = surgery))
p3_3 <- p3_3 + geom_point() +
  geom_line()

p4_3 <- ggplot(data = data3, aes(x = time, y = ip.10, group = horse, color = surgery))
p4_3 <- p4_3 + geom_point() +
  geom_line()

p5_3 <- ggarrange(p1_3, p2_3, p3_3, p4_3, ncol = 2, nrow = 2, common.legend = TRUE, legend = "bottom")

p5_3
```

## Interindividual distribution looks better now - run models on all 4 cytokines using surgery (Y/N), time, and an interaction term with individual horse as a random effect.

```{r, echo=FALSE}
model3_il.18 <- lmer(il.18 ~ surgery*time + (1|horse), data = data3)
summary(model3_il.18)

model3_il.2 <- lmer(il.2 ~ surgery*time + (1|horse), data = data3)
summary(model3_il.2)

model3_ip.10 <- lmer(ip.10 ~ surgery*time + (1|horse), data = data3)
summary(model3_ip.10)

model3_il.10 <- lmer(il.10 ~ surgery*time + (1|horse), data = data3)
summary(model3_il.10)
```

## Focus on IL-10, as this is the only one that demonstrated a difference between horses that had surgery and those that didn't.
## Distribution is not great, so try log transformation of all 4 ###

```{r}
data4 <- data3 %>% mutate(log_il.18 = if_else(il.18 <= 0, 0, log(il.18)),
                          log_il.2 = if_else(il.2 <= 0, 0, log(il.2)),
                          log_ip.10 = if_else(ip.10 <= 0, 0, log(ip.10)),
                          log_il.10 = if_else(il.10 <= 0, 0, log(il.10)))

log_cytokine_vars <- list('log_il.18', 'log_il.2', 'log_ip.10', 'log_il.10')
for (i in log_cytokine_vars){
  model <- lmer(paste(i[[1]], "~ surgery*time + (1|horse)"), data = data4)
  print(anova(model))
  plot(resid(model, type = "pearson") ~ fitted(model))
  qqnorm(resid(model, type = "pearson"))
}
### IL10 seems to be the one significant factor here ###
```

## Focus on IL-10 and run contrasts

```{r}
model4_il.10 <- lmer(log_il.10 ~ surgery*factor(time) + (1|horse), data = data4) #factor time to allow for individual comparisons
### Check contrasts
emm_model1 <- emmeans(model4_il.10, ~surgery|time, at = list(time = c(0, 24, 48)))
contrast(emm_model1, "trt.vs.ctrl")
emm_model2 <- emmeans(model4_il.10, ~time|surgery, at = list(time = c(0, 24, 48)))
contrast(emm_model2, "pairwise")
```

# IL-10 is significantly different in the first 48 hours when comparing horses that had surgery (lower) to those that didn't (higher), although there is no significant change over time in these 2 groups.

```{r}
p2_3_final <- ggplot(data = data3, aes(x = time, y = il.10))
p2_3_final <- p2_3_final + geom_point(aes(shape = surgery, group = horse)) +
  geom_line(aes(linetype = surgery, group = horse)) +
  stat_summary(aes(group = surgery, shape = surgery), geom = "point", position = position_dodge(width = 2), size = 4) +
  stat_summary(aes(group = surgery), linetype = 'solid', geom = "line") +
  stat_summary(aes(group = surgery), geom = "errorbar", width = 2, position = position_dodge(width = 2)) +
  scale_linetype_manual(values = c(2,3)) +
  scale_x_continuous(breaks = c(0, 24, 48)) +
  theme_bw()
p2_3_final
```

## Graph physiologic/CBC variables
```{r}
### Plot other data ###

p1_4 <- ggplot(data = data, aes(x = time, y = temp, group = horse, color = surgery))
p1_4 <- p1_4 + geom_point() +
  geom_line()

p2_4 <- ggplot(data = data, aes(x = time, y = wbc, group = horse, color = surgery))
p2_4 <- p2_4 + geom_point() +
  geom_line()

p3_4 <- ggplot(data = data, aes(x = time, y = neut, group = horse, color = surgery))
p3_4 <- p3_4 + geom_point() +
  geom_line()

p4_4 <- ggplot(data = data, aes(x = time, y = lymph, group = horse, color = surgery))
p4_4 <- p4_4 + geom_point() +
  geom_line()

p5_4 <- ggplot(data = data, aes(x = time, y = fibr, group = horse, color = surgery))
p5_4 <- p5_4 + geom_point(position = position_dodge(width = 2)) +
  geom_line()

p6_4 <- ggarrange(p1_4, p2_4, p3_4, p4_4, p5_4, ncol = 3, nrow = 2, common.legend = TRUE, legend = "bottom")

p6_4
```

## Nothing stands out, so model individually
```{r}
phys_vars <- list('temp', 'wbc', 'neut', 'lymph', 'fibr')
for (i in phys_vars){
  model <- lmer(paste(i[[1]], "~ surgery*time + (1|horse)"), data = data)
  print(anova(model))
  plot(resid(model, type = "pearson") ~ fitted(model))
  qqnorm(resid(model, type = "pearson"))
}
```

## Looks like neutrophils and lymphocytes have some significant differences
## resids look fine, let's check the contrasts for both
### First neutrophils
```{r}
model_neut <- lmer(neut ~ surgery*factor(time) + (1|horse), data = data) #factor time to allow for individual comparisons
### Check contrasts ###
emmeans(model_neut, ~surgery|time, at = list(time = c(0, 24, 48))) %>%
  contrast("trt.vs.ctrl")
emmeans(model_neut, ~time|surgery, at = list(time = c(0, 24, 48))) %>%
  contrast("pairwise")
```
## Then lymphocytes
```{r}
model_lymph <- lmer(lymph ~ surgery*factor(time) + (1|horse), data = data) #factor time to allow for individual comparisons
### Check contrasts ###
emmeans(model_lymph, ~surgery|time, at = list(time = c(0, 24, 48))) %>%
  contrast("trt.vs.ctrl")
emmeans(model_lymph, ~time|surgery, at = list(time = c(0, 24, 48))) %>%
  contrast("pairwise")
```
## And graphically
```{r}
### Neutrophil comparison ###
p_neut_final <- ggplot(data = data, aes(x = time, y = neut))
p_neut_final <- p_neut_final + geom_point(aes(shape = surgery, group = horse)) +
  geom_line(aes(linetype = surgery, group = horse)) +
  stat_summary(aes(group = surgery, shape = surgery), geom = "point", position = position_dodge(width = 2), size = 4) +
  stat_summary(aes(group = surgery), linetype = 'solid', geom = "line") +
  stat_summary(aes(group = surgery), geom = "errorbar", width = 2, position = position_dodge(width = 2)) +
  scale_linetype_manual(values = c(2,3)) +
  scale_x_continuous(breaks = c(0, 24, 48)) +
  theme_bw()
p_neut_final

### Lymphocyte comparison ###
p_lymph_final <- ggplot(data = data, aes(x = time, y = lymph))
p_lymph_final <- p_lymph_final + geom_point(aes(shape = surgery, group = horse)) +
  geom_line(aes(linetype = surgery, group = horse)) +
  stat_summary(aes(group = surgery, shape = surgery), geom = "point", position = position_dodge(width = 2), size = 4) +
  stat_summary(aes(group = surgery), linetype = 'solid', geom = "line") +
  stat_summary(aes(group = surgery), geom = "errorbar", width = 2, position = position_dodge(width = 2)) +
  scale_linetype_manual(values = c(2,3)) +
  scale_x_continuous(breaks = c(0, 24, 48)) +
  theme_bw()
p_lymph_final
```
# There was a significant effect of surgery on neutrophil counts, which were increased at all time points in the surgery group, but did not change from baseline in either group
# Lymphocyte demonstrated a significanntly different change over time (the slopes of the best fit lines are different), with counts decreased at 24 and 48 hours in the surgery group over the non-surgery group, but no differences at baseline

