---
title: "Postop Cytokines over Time"
author: "Gregg Griffenhagen"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(emmeans)
library(lme4)
library(lmerTest)
library(ggplot2)
library(ggpubr)
data <- read.csv('postop_cytok_over_time.csv')
data$surgery <- as.factor(data$surgery)
data$horse <- as.factor(data$horse)
```

## Quick Data Summary

```{r, warning=FALSE}
data %>% group_by(time, surgery) %>%
  summarise(across(c(il.18, il.2, ip.10, il.10),
                   c(mean = ~mean(.x), sd = ~sd(.x)))) %>%
  kableExtra::kable()
```

## Nothing is significant, but the plots are interesting
# Start with the full data set

```{r, echo=FALSE}
p1 <- ggplot(data = data, aes(x = time, y = il.18, group = horse, color = surgery))
p1 <- p1 + geom_point() +
  geom_line()

p2 <- ggplot(data = data, aes(x = time, y = il.10, group = horse, color = surgery))
p2 <- p2 + geom_point() +
  geom_line()

p3 <- ggplot(data = data, aes(x = time, y = il.2, group = horse, color = surgery))
p3 <- p3 + geom_point() +
  geom_line()

p4 <- ggplot(data = data, aes(x = time, y = ip.10, group = horse, color = surgery))
p4 <- p4 + geom_point() +
  geom_line()

p5 <- ggarrange(p1, p2, p3, p4, ncol = 2, nrow = 2, common.legend = TRUE, legend = "bottom")

p5
```

## Remove the obvious outlier at the top and recheck plots

```{r, echo=FALSE}
data2 <- data %>% filter(horse != 13) #removes the outlier manually

p1_2 <- ggplot(data = data2, aes(x = time, y = il.18, group = horse, color = surgery))
p1_2 <- p1_2 + geom_point() +
  geom_line()

p2_2 <- ggplot(data = data2, aes(x = time, y = il.10, group = horse, color = surgery))
p2_2 <- p2_2 + geom_point() +
  geom_line()

p3_2 <- ggplot(data = data2, aes(x = time, y = il.2, group = horse, color = surgery))
p3_2 <- p3_2 + geom_point() +
  geom_line()

p4_2 <- ggplot(data = data2, aes(x = time, y = ip.10, group = horse, color = surgery))
p4_2 <- p4_2 + geom_point() +
  geom_line()

p5_2 <- ggarrange(p1_2, p2_2, p3_2, p4_2, ncol = 2, nrow = 2, common.legend = TRUE, legend = "bottom")

p5_2
```

## The 3 interleukins still demonstrate a single horse as an outlier (although not as bad)
## Let's remove that one too

```{r, echo=FALSE}
data3 <- data2 %>% filter(horse != 2) # again remove the outlier manually

p1_3 <- ggplot(data = data3, aes(x = time, y = il.18, group = horse, color = surgery))
p1_3 <- p1_3 + geom_point() +
  geom_line()

#FOCUS ON THIS ONE  - seems to be the only one with differences
p2_3 <- ggplot(data = data3, aes(x = time, y = il.10, group = horse, color = surgery))
p2_3 <- p2_3 + geom_point() +
  geom_line()

p3_3 <- ggplot(data = data3, aes(x = time, y = il.2, group = horse, color = surgery))
p3_3 <- p3_3 + geom_point() +
  geom_line()

p4_3 <- ggplot(data = data3, aes(x = time, y = ip.10, group = horse, color = surgery))
p4_3 <- p4_3 + geom_point() +
  geom_line()

p5_3 <- ggarrange(p1_3, p2_3, p3_3, p4_3, ncol = 2, nrow = 2, common.legend = TRUE, legend = "bottom")

p5_3
```

## Interindividual distribution looks better now - run models on all 4 cytokines using surgery (Y/N), time, and an interaction term with individual horse as a random effect.

```{r}
model3_il.18 <- lmer(il.18 ~ surgery*time + (1|horse), data = data3)
summary(model3_il.18)

model3_il.2 <- lmer(il.2 ~ surgery*time + (1|horse), data = data3)
summary(model3_il.2)

model3_ip.10 <- lmer(ip.10 ~ surgery*time + (1|horse), data = data3)
summary(model3_ip.10)

model3_il.10 <- lmer(il.10 ~ surgery*time + (1|horse), data = data3)
summary(model3_il.10)
```

## None of the interactions are significant (no change in the time effect depending on whether surgery or not), so re-run without interaction term

```{r}
model3_il.18 <- lmer(il.18 ~ surgery+time + (1|horse), data = data3)
summary(model3_il.18)

model3_il.2 <- lmer(il.2 ~ surgery+time + (1|horse), data = data3)
summary(model3_il.2)

model3_ip.10 <- lmer(ip.10 ~ surgery+time + (1|horse), data = data3)
summary(model3_ip.10)

model3_il.10 <- lmer(il.10 ~ surgery+time + (1|horse), data = data3)
summary(model3_il.10)
```

## Focus on IL-10, as this is the only one that demonstrated a difference between horses that had surgery and those that didn't.
## Need to check the distribution of the data

```{r}
shapiro.test(data3$il.10)
plot_norm1 <- plot(model3_il.10)
plot_norm2 <- ggplot(data = data3, aes(x = il.10)) +
  geom_histogram()
plot_norm_1 <- ggarrange(plot_norm1, plot_norm2, nrow = 1)
plot_norm_1
```

## So not distributed super well - try log transforming without the outlier data

```{r}
data4 <- data3 %>% mutate(log_il.10 = if_else(il.10 <= 0, 0, log(il.10)))

model4_il.10 <- lmer(log_il.10 ~ surgery+time + (1|horse), data = data4)

shapiro.test(data4$log_il.10)
plot_norm3 <- plot(model4_il.10)
plot_norm4 <- ggplot(data = data4, aes(x = log_il.10)) +
  geom_histogram()
plot_norm_2 <- ggarrange(plot_norm3, plot_norm4, nrow = 1)
plot_norm_2
```

## Slightly better - should probably use this transformation for analysis
```{r}
p2_3
anova(model4_il.10)
```

# IL-10 is significantly different in the first 48 hours when comparing horses that had surgery (lower) to those that didn't (higher), although there is no significant change over time in these 2 groups.



